%%
% This is an Overleaf template for scientific articles and reports
% using the TUM Corporate Desing https://www.tum.de/cd
%
% For further details on how to use the template, take a look at our
% GitLab repository and browse through our test documents
% https://gitlab.lrz.de/latex4ei/tum-templates.
%
% The tumarticle class is based on the KOMA-Script class scrartcl.
% If you need further customization please consult the KOMA-Script guide
% https://ctan.org/pkg/koma-script.
% Additional class options are passed down to the base class.
%
% If you encounter any bugs or undesired behaviour, please raise an issue
% in our GitLab repository
% https://gitlab.lrz.de/latex4ei/tum-templates/issues
% and provide a description and minimal working example of your problem.
%%


\documentclass[
  english,        % define the document language (english, german)
  font=times,     % define main text font (helvet, times, palatino, libertine)
  twocolumn,      % use onecolumn or twocolumn layout
]{tumarticle}


% load additional packages
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{url}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, fit}
\usepackage{forest}
\usepackage{adjustbox}

\setlength\parindent{0pt}


% article metadata
\title{Agentic AI for Code Generation}
\subtitle{Seminar Report}

\author[affil=1, email=thuaduc.nguyen@tum.de]{Student: Thua Duc Nguyen}

\author[affil=1, email=derui.zhu@tum.de]{Supervisor: Derui Zhu}


\affil[mark=1]{\theDepartmentName, \theUniversityName}



\date{June 30, 2026}


\begin{document}

\maketitle

\begin{abstract}
  Large language models have moved from autocompleting single functions to
  autonomously resolving real software issues end to end, a shift popularly
  termed ``vibe coding.'' This report surveys autonomous AI agents for code
  generation through a single unifying lens: a five-stage \emph{vibe coding
  pipeline} (intent capture, planning, code generation, execution, and
  refinement). Using this framing we synthesise three active research fronts.
  First, we survey end-to-end software-engineering agents (SWE-agent, Agentless,
  AutoCodeRover, OpenHands, Devin) and the architectural patterns they share: the
  observe--think--act loop, agent--computer interfaces, code-as-action, and
  repository-scale context management, together with their evaluation on
  SWE-bench and its variants. Second, we examine planning and search strategies,
  from chain/tree/graph-of-thought reasoning to hierarchical localisation and
  sample-and-rank candidate selection. Third, we analyse execution feedback and
  iterative refinement, covering feedback signals, self-repair mechanisms,
  fault localisation, and convergence criteria. Across all three fronts we find
  recurring patterns: feedback-driven loops, execution-grounded selection, and a
  persistent context bottleneck, and we highlight open challenges in evaluation
  realism, cost, and trust. The pipeline framing exposes how planning and
  refinement, often studied in isolation, function as a coupled control layer
  over generation, but only when an \emph{external} oracle closes the loop.
  Internal self-critique, we argue, is an open loop wearing a control loop's
  clothes; and intent capture, alone among the five stages, has no feedback
  mechanism at all.
\end{abstract}

\section{Introduction}

\paragraph{Motivation.}
Coding with a language model used to mean accepting an autocomplete suggestion.
Increasingly it means describing a change in plain English and letting a system
find the right files, edit them, run the tests, and iterate until they pass.
The popular name for this style, ``vibe coding''~\cite{karpathy2025vibecoding},
captures both its appeal and
the worry it raises: the developer states an intent and reviews an outcome,
while the steps in between (reading the codebase, forming a hypothesis, running
the debugger) are delegated to the model. The measured progress is real. In
late~2023 the best system on SWE-bench, a benchmark of real GitHub issues,
resolved under 2\% of them~\cite{jimenez2024swebench}; two years later agents
resolve most of the human-verified split. This report asks what produced that
jump, and what it still does not deliver.

\paragraph{Background and Related Surveys.}
Existing surveys catalogue agents in software engineering~\cite{wang2024agentsse}
and LLM-based code-generation agents~\cite{dong2025surveycodegen}, organised
largely by system or by capability. This report instead organises three active
research fronts (end-to-end agents, planning and search, and execution-feedback
refinement) around the pipeline stages they instantiate, which lets us compare
systems that describe themselves in incompatible vocabularies.

\paragraph{Scope and Contribution.}
We cover agents whose output is a code change: repository-level issue
resolution, automated program repair, and function-level synthesis, drawing on
the 2023--2026 literature. Editor autocompletion, code search, and multi-agent
role-play frameworks lie outside this scope. The report contributes, first, the
vibe coding pipeline as a shared vocabulary for these systems
(\S\ref{sec:pipeline}); second, a survey of the three fronts read against that
pipeline (\S\ref{sec:e2e}--\S\ref{sec:refinement}); and third, a consolidation
of what the fronts share and where they are jointly stuck
(\S\ref{sec:discussion}), drawing on 2025--2026 evidence about benchmark
contamination, cost, and security that postdates most existing surveys.

\section{The Vibe Coding Pipeline}
\label{sec:pipeline}

\subsection{What ``Vibe Coding'' Means Here}
\label{sec:vibe-definition}

The term is Karpathy's, coined in February~2025 to describe a way of working in
which one ``fully give[s] in to the vibes\ldots{} and forget[s] that the code
even exists''~\cite{karpathy2025vibecoding}. In that original sense vibe coding
names a \emph{human posture toward the output}: the developer states an intent,
accepts the diff without reading it, and iterates on observed behaviour rather
than on the source. It is a claim about what the human stops doing, not about
how the machine is built.

We use the term in a deliberately broader, system-side sense. The \emph{vibe
coding pipeline} is the machinery that makes that posture available: the
sequence of stages by which an agent turns an unread intent into a change it has
itself validated. The generalisation is worth stating explicitly because it
carries an obligation. If the developer no longer reads the diff, every
guarantee the developer used to supply must be re-supplied by some stage of the
pipeline. Which stages can actually supply such a guarantee, and which only
appear to, is the question this report pursues. \S\ref{sec:open} and
\S\ref{sec:practice} return to Karpathy's narrow sense to argue that
verification is precisely the guarantee the pipeline does \emph{not} yet
re-supply.

\subsection{Five Stages}

Papers on autonomous coding agents describe their systems in their own terms:
phases, modules, roles, loops. Underneath, nearly all of them do the same five
things. Naming those five stages gives us a frame for comparing systems that
otherwise share no vocabulary. The pipeline is not a new architecture but a
lens on existing ones.

Its stages echo the classical waterfall phases (requirements, design,
implementation, testing), yet the resemblance is superficial. An agent traverses
all five in seconds and repeats them many times per task, and the loop is closed
by a test runner rather than by a review meeting. The closer analogue is the
developer's inner loop of edit, run, and debug, with planning and code
localisation folded into it.

\begin{figure}[t]
  \centering
  \resizebox{\columnwidth}{!}{%
  \begin{tikzpicture}[
    stage/.style={draw, rounded corners=3pt, fill=blue!12, font=\small,
                  align=center, minimum height=0.9cm, minimum width=2.4cm},
    note/.style={font=\scriptsize, align=center, inner sep=1pt},
    arr/.style={-{Stealth[length=1.8mm]}, very thick},
    feedback/.style={-{Stealth[length=1.8mm]}, very thick, dashed, red!60},
    fblab/.style={font=\tiny, text=red!70, align=center, inner sep=1.5pt,
                  fill=white},
  ]
    \node[stage] (intent) at (0,0)      {Intent\\Capture};
    \node[stage] (plan)   at (3.6,0)    {Planning};
    \node[stage] (gen)    at (7.2,0)    {Code\\Generation};
    \node[stage] (exec)   at (7.2,-2.4) {Execution};
    \node[stage] (refine) at (3.6,-2.4) {Refinement};

    \draw[arr] (intent) -- (plan);
    \draw[arr] (plan) -- (gen);
    \draw[arr] (gen) -- (exec);
    \draw[arr] (exec) -- (refine);
    \draw[arr] (refine.west) to[out=180, in=270] (intent.south);

    \draw[feedback] (refine.north) -- node[fblab, right] {replan} (plan.south);
    \draw[feedback] (refine.north east) -- node[fblab, pos=0.5] {regenerate} (gen.south west);

    \node[note, above=0.2cm of intent] {NL issue or spec\\\S\ref{sec:intent}};
    \node[note, above=0.2cm of plan]   {\S\ref{sec:planning}};
    \node[note, below=0.2cm of refine] {\S\ref{sec:refinement}};
    \node[note, below=0.2cm of exec]   {test failures,\\traces};
  \end{tikzpicture}%
  }
  \caption{The vibe coding pipeline. Intent flows through planning, generation,
    and execution. Execution failures drive refinement, which either regenerates
    the patch or replans when the initial hypothesis is invalidated.}
  \label{fig:vibe-pipeline}
\end{figure}

The vibe coding pipeline (Figure~\ref{fig:vibe-pipeline}) consists of five
stages forming a loop: Intent Capture \textrightarrow{} Planning
\textrightarrow{} Code Generation \textrightarrow{} Execution
\textrightarrow{} Refinement. Execution feedback drives refinement, which may
trigger re-planning or re-generation. Adjacent stages exchange structured
artefacts: natural-language intent and repository context flow into plans and
candidate patches; generated code is run against test oracles; failures and
traces feed back into replanning or repair.

\subsection{Mapping to Survey Topics}
Each of the three main sections corresponds to one or more pipeline stages.
\S\ref{sec:e2e} covers the full pipeline as an integrated system, including
action interfaces for code generation (ACI, code-as-action, diff-based edits);
\S\ref{sec:planning} treats intent capture (\S\ref{sec:intent}) together with
planning, candidate search, and the plan--generate interface, since in practice
an agent operationalises intent and plans over it in the same breath;
\S\ref{sec:refinement} covers execution environments, test-driven feedback, and
the refinement loop (including fault-specific self-repair). Code generation is
deliberately not given its own section: no surveyed system treats emitting
tokens as a separable problem, and the design decisions that govern it (action
space, edit format, candidate sampling) belong to the stages on either side of
it. It is therefore distributed across \S\ref{sec:e2e} and
\S\ref{sec:planning}. Cross-cutting concerns (evaluation, context management)
are consolidated in the Discussion.
Figure~\ref{fig:survey-taxonomy} provides a document-wide taxonomy: report
sections mapped to pipeline stages, with representative systems and techniques
as leaf exemplars.

\begin{figure*}[tp]
  \centering
  \begin{adjustbox}{max size={\textwidth}{\textheight}}
  \begin{forest}
    for tree={
      rounded corners,
      draw,
      align=center,
      font=\small,
      edge={-stealth, semithick},
      l sep=6mm,
      s sep=1.6mm,
      anchor=west,
      child anchor=west,
      parent anchor=east,
      grow'=east,
      fill=gray!10,
      text width=8.5em,
      leaf/.style={font=\footnotesize, text width=24em},
    }
    [{Agentic AI for\\Code Generation}, text width=9.5em, fill=orange!10
      [{End-to-End\\Systems (\S\ref{sec:e2e})}, text width=8em
        [{Orchestration}
          [{Agentic observe--think--act loop\\(SWE-agent, OpenHands, Devin)}, leaf, fill=blue!5]
          [{Fixed pipeline (Agentless)}, leaf, fill=blue!5]
        ]
        [{Action\\interface}
          [{Custom ACI; code-as-action;\\AST APIs; diff edits}, leaf, fill=blue!5]
        ]
        [{Context\\management}
          [{Compression; retrieval;\\AST+SBFL; sandbox; memory}, leaf, fill=blue!5]
        ]
        [{Evaluation\\(\S\ref{sec:e2e-eval})}
          [{SWE-bench Full/Lite/Verified;\\difficulty skew; leaderboard trajectory}, leaf, fill=blue!5]
        ]
      ]
      [{Planning \&\\Search (\S\ref{sec:planning})}, text width=8em
        [{Intent capture\\(\S\ref{sec:intent})}
          [{Issue quality; underspecification;\\open-loop stage, no clarification}, leaf, fill=green!5]
        ]
        [{Plan\\representations}
          [{CoT, ToT, GoT; specification sketches;\\localisation plans (Agentless, AutoCodeRover)}, leaf, fill=green!5]
        ]
        [{Search\\strategies}
          [{Self-consistency; repo navigation;\\sample-and-rank (AlphaCode, CodeT)}, leaf, fill=green!5]
        ]
        [{Plan--generate\\interface}
          [{Plan-then-generate; interleaved (ReAct);\\generate-then-refine (Self-Refine)}, leaf, fill=green!5]
        ]
      ]
      [{Execution\\Feedback (\S\ref{sec:refinement})}, text width=8em
        [{Feedback\\signals}
          [{Tests, compiler errors, static analysis;\\external vs.\ internal; sandboxes}, leaf, fill=red!5]
        ]
        [{Refinement\\mechanisms}
          [{Reflexion; self-debugging;\\SelfEvolve; diff-based repair}, leaf, fill=red!5]
        ]
        [{Limits of\\self-correction (\S\ref{sec:refinement-limits})}
          [{Budget-normalised self-repair;\\intrinsic self-correction fails}, leaf, fill=red!5]
        ]
        [{Fault\\localisation}
          [{SBFL; FL-free repair (D4C);\\stack-trace reasoning; ranking \& ensembles}, leaf, fill=red!5]
        ]
        [{Convergence\\(\S\ref{sec:convergence})}
          [{Pass thresholds; diminishing returns;\\iteration/token budgets; overfitting; HITL}, leaf, fill=red!5]
        ]
      ]
      [{Discussion\\(\S\ref{sec:discussion})}, text width=8em
        [{Consolidated\\patterns}
          [{Observe--think--act loop; execution-grounded selection;\\planning as control layer; inference-time scaling}, leaf, fill=yellow!8]
        ]
        [{Evaluation\\landscape}
          [{Saturation (HumanEval, MBPP); SWE-bench Pro \& Multilingual;\\contamination, reward hacking; SWE-Effi cost-aware scoring}, leaf, fill=yellow!8]
        ]
        [{Context\\bottleneck}
          [{Dependency-aware retrieval; context explosion \&\\semantic drift; active context management}, leaf, fill=yellow!8]
        ]
        [{Open\\challenges}
          [{Long-horizon multi-file reasoning; specification\\ambiguity; cost; security \& trust}, leaf, fill=yellow!8]
        ]
      ]
    ]
  \end{forest}%
  \end{adjustbox}
  \caption{Survey taxonomy organised by report section and vibe coding pipeline
    stage. \S\ref{sec:e2e} surveys integrated end-to-end systems;
    \S\ref{sec:planning} and \S\ref{sec:refinement} zoom into planning/search
    and execution-feedback mechanisms; \S\ref{sec:discussion} consolidates the
    recurring patterns, the evaluation landscape, the context bottleneck, and
    the open challenges. Leaf nodes name representative systems or techniques.}
  \label{fig:survey-taxonomy}
\end{figure*}

\section{End-to-End Autonomous Software Engineering Agents}
\label{sec:e2e}

Before examining individual pipeline stages, we survey systems that orchestrate
the \emph{entire} vibe coding pipeline, accepting a natural-language issue or
specification and producing a tested code change with minimal human
intervention. These end-to-end agents are the setting in which
planning (\S\ref{sec:planning}) and feedback-driven refinement
(\S\ref{sec:refinement}) operate.


We begin by identifying the architectural patterns that underlie
most current systems (\S\ref{sec:e2e-arch}). We then survey five representative
agents that span the design space: from fully autonomous to deliberately
non-agentic (\S\ref{sec:e2e-systems}). Finally, we examine how these systems
are evaluated, focusing on SWE-bench and its variants
(\S\ref{sec:e2e-eval}). This section instantiates the \emph{End-to-End Systems}
branch of Figure~\ref{fig:survey-taxonomy}.

\begin{table*}[t]
  \centering
  \caption{Comparison of end-to-end autonomous SE agents. Architectural
    patterns: \textbf{OTA}~=~observe--think--act loop,
    \textbf{ACI}~=~custom Agent--Computer Interface,
    \textbf{CaA}~=~code-as-action,
    \textbf{Pipe}~=~fixed multi-phase pipeline (no agent loop).
    Splits are defined in Table~\ref{tab:swebench-splits}; all scores are
    single-pass (pass@1) on each system's base configuration, so ablations and
    pass@$k$ variants reported in the text are higher.
    Cost is the average USD per instance on Lite at pass@1. All figures are
    as-published by each system's own authors and reflect the backbone model of
    that publication, not current frontier models.}
  \label{tab:e2e-systems}
  \small
  \setlength{\tabcolsep}{3.5pt}
  \begin{tabular}{@{}llllccc@{}}
    \toprule
    \textbf{System} & \textbf{Architecture} & \textbf{Backbone}
      & \textbf{Context Strategy} & \textbf{Lite} & \textbf{Full} & \textbf{Cost} \\
    \midrule
    SWE-agent~\cite{yang2024sweagent} & OTA + ACI & GPT-4 Turbo
      & Observation compression & 18.0\% & 12.5\% & \$1.67 \\
    Agentless~\cite{xia2024agentless} & Pipe & GPT-4o
      & Repo skeleton + embeddings & 32.0\% & --$^{\dagger}$ & \$0.70 \\
    AutoCodeRover~\cite{zhang2024autocoderover} & OTA + AST search & GPT-4
      & AST APIs (SBFL optional) & 19.0\% & 12.4\% & \$0.43 \\
    OpenHands~\cite{wang2024openhands} & OTA + CaA & Claude 3.5 Sonnet
      & Docker sandbox + delegation & 26.0\% & --$^{\dagger}$ & \$1.10 \\
    Devin~\cite{cognition2024devin} & OTA + Brain/DevBox & undisclosed
      & Persistent knowledge store & -- & 13.9\%$^{\ddagger}$ & --$^{\ddagger}$ \\
    \bottomrule
    \multicolumn{7}{@{}p{\textwidth}@{}}{\scriptsize $^{\dagger}$Not evaluated on Full; the authors cite compute cost. \quad $^{\ddagger}$Devin is closed-source: the backbone model and per-instance cost are undisclosed, and the resolve rate is self-reported on a random 25\% subset of Full. No peer-reviewed paper exists and the result has not been independently reproduced.}
  \end{tabular}
\end{table*}

\subsection{Architecture and Orchestration}
\label{sec:e2e-arch}

Despite surface-level diversity, autonomous SE agents converge on a small
number of architectural patterns. As illustrated in the
\emph{Orchestration}, \emph{Action Interface}, and \emph{Context Management}
branches of Figure~\ref{fig:survey-taxonomy}, these patterns determine how agents
perceive their environment, what actions they can take, and how they manage
repository-scale context.

\paragraph{The observe--think--act loop.}
The dominant paradigm is a ReAct-style~\cite{yao2023react} loop in which the
agent alternates between (i)~observing the environment (file contents, test
output, error traces), (ii)~reasoning about what to do next (chain-of-thought
or structured plan), and (iii)~executing an action via a tool call. SWE-agent~\cite{yang2024sweagent} and OpenHands~\cite{wang2024openhands}
both instantiate this loop, differing mainly in the \emph{action space} exposed
to the model (Figure~\ref{fig:react-loop}).

\begin{figure}[t]
  \centering
  \resizebox{\columnwidth}{!}{%
  \begin{tikzpicture}[
    node distance=0.35cm,
    box/.style={
      draw,
      rounded corners=2pt,
      align=center,
      font=\scriptsize,
      inner sep=2pt,
      minimum height=0.45cm,
    },
    env/.style={box, fill=gray!12, minimum width=3.6cm},
    phase/.style={box, fill=blue!6, minimum width=2.0cm},
    arr/.style={-{Stealth[length=1.4mm]}, semithick},
  ]
    \node[env] (env) {Environment\\\tiny files, tests, errors};
    \node[phase, below=of env] (observe) {(i)~Observe};
    \node[phase, below=of observe] (think) {(ii)~Think};
    \node[phase, below=of think] (act) {(iii)~Act};

    \draw[arr] (observe.south) -- (think.north);
    \draw[arr] (think.south) -- (act.north);

    \coordinate (toolOut) at ([xshift=6mm]env.east);
    \draw[arr] (act.east) -- (act.east -| toolOut)
      -- node[pos=0.35, right, font=\tiny, align=left] {tool\\call} (toolOut)
      -- (env.east);

    \coordinate (stateOut) at ([xshift=-6mm]env.west);
    \draw[arr] (env.west) -- (stateOut)
      -- node[pos=0.35, left, font=\tiny, align=right] {updated\\state} (stateOut |- observe.west)
      -- (observe.west);
  \end{tikzpicture}%
  }
  \caption{ReAct-style observe--think--act loop. The agent reads environment
    state, reasons about the next step, executes a tool call, and repeats.}
  \label{fig:react-loop}
\end{figure}

\paragraph{Agent--Computer Interface (ACI).}
Yang et al.~\cite{yang2024sweagent} introduce the ACI concept, arguing that
LLMs are a ``new category of end users'' that need purpose-built interfaces
rather than raw shell access. SWE-agent's ACI includes a windowed file viewer
(100~lines), a linter-guarded \texttt{edit} command, summarised search results,
and compressed observation history. Ablations show that each ACI component
contributes measurably: removing the linter drops performance by 3~percentage
points; switching to iterative search costs 6~pp; and showing full file content
instead of a 100-line window costs 5.3~pp on SWE-bench Lite.

\paragraph{Code-as-action.}
OpenHands adopts the CodeAct paradigm~\cite{wang2024codeact}, consolidating all
agent actions into executable Python and bash code rather than a fixed tool
API. Each turn may run arbitrary IPython cells, bash commands, or browser
interactions inside a Docker-sandboxed environment. This design maximises
flexibility at the cost of a larger action space that the model must navigate.

\paragraph{Context management.}
Repository-level tasks easily exceed LLM context windows. Agents employ several
mitigation strategies: SWE-agent collapses older observations beyond the last
5~turns; AutoCodeRover~\cite{zhang2024autocoderover} parses the codebase into
an AST and exposes structure-aware search APIs (\texttt{search\_class},
\texttt{search\_method\_in\_class}) that return only signatures, not full
definitions; Agentless~\cite{xia2024agentless} constructs a hierarchical
repository skeleton and uses embedding-based retrieval to select candidate
files before prompting the LLM.

\subsection{Representative Systems}
\label{sec:e2e-systems}

We survey five prominent systems that span the design space.
Table~\ref{tab:e2e-systems} provides a comparative overview.

\paragraph{SWE-agent~\cite{yang2024sweagent}.}
Developed at Princeton, SWE-agent wraps a GPT-4 Turbo backbone in a
custom ACI with specialised file navigation, editing, and search commands.
It achieved 12.47\% on the full SWE-bench test set (2{,}294 tasks) and
18.00\% on SWE-bench Lite when first published. The agent follows a
typical trajectory: reproduce the issue, localise the faulty code via
\texttt{find\_file}/\texttt{search\_dir}, then enter an edit--execute loop.
Runs average \$1.67 per instance on Lite, but the distribution is bimodal:
successful trajectories finish in a median of 12~steps at \$1.21, whereas
failures burn a mean of 21~steps and \$2.52 before giving up.

\paragraph{Agentless~\cite{xia2024agentless}.}
From UIUC, Agentless challenges the need for agentic autonomy. It replaces
the interactive loop with a fixed three-phase pipeline:
\emph{localise}~$\to$~\emph{repair}~$\to$~\emph{validate}. Localisation uses
a hierarchical narrowing strategy (files~$\to$~classes/functions~$\to$~edit
locations) combining LLM prompting with embedding retrieval. Repair samples
42~candidate patches per bug in a search/replace diff format. Validation
selects the best patch via regression testing, LLM-generated reproduction
tests, and majority voting over AST-normalised candidates. With a GPT-4o
backbone, Agentless resolved 32.00\% of SWE-bench Lite at an average cost of
only \$0.70 per problem, the highest reported figure among open-source
systems at the time and a strong non-agentic baseline. OpenAI subsequently
adopted Agentless as the scaffold for reporting GPT-4o and o1 results on
SWE-bench Verified, describing it as the best-performing open-source scaffold
available~\cite{openai2024verified}.

\paragraph{AutoCodeRover~\cite{zhang2024autocoderover}.}
From NUS, AutoCodeRover takes a ``software-engineering-oriented'' approach,
operating on program representations rather than raw files. It parses the
codebase into an AST and provides the LLM with structure-aware search APIs.
A stratified context retrieval process iteratively expands the agent's
understanding, starting from issue keywords and refining through multiple
search rounds until sufficient context is gathered. When a test suite is
available, Spectrum-Based Fault Localisation (SBFL) augments the search with
suspiciousness scores. In its base configuration on a GPT-4 backbone,
AutoCodeRover resolved 19.0\% of Lite and 12.4\% of Full in a single pass---the
figures reported in Table~\ref{tab:e2e-systems}. Two extensions raise this
without changing the backbone: enabling SBFL lifts the Lite rate to 22\%, uniquely
solving 7~instances no other configuration reaches, while counting the union of
three sampled runs (pass@3) rather than one raises the base figures to 26.0\% on
Lite and 18.0\% on Full. Cost ranges from \$0.43 per task at pass@1 to \$1.30 at
pass@3.

\paragraph{OpenHands~\cite{wang2024openhands}.}
Formerly OpenDevin, OpenHands is an open-source platform (ICLR~2025,
32K~GitHub stars, 188~contributors) built around the CodeAct agent. Each
session runs inside a Docker container with a bash shell, IPython server, and
Chromium browser. The platform supports multi-agent delegation: a generalist
CodeAct agent can offload web browsing to a specialised BrowsingAgent. On
SWE-bench Lite, CodeAct~v1.8 with Claude~3.5~Sonnet achieved 26.0\%.
On SWE-bench Verified, a later submission reaches 60.6\% with a single
rollout and 66.4\% when five trajectories are reranked by a dedicated critic
model~\cite{openhands2025verified}.

\paragraph{Devin (Cognition)~\cite{cognition2024devin}.}
Devin is a proprietary cloud-hosted agent positioned as an ``autonomous AI
software engineer.'' Its architecture separates a \emph{Brain} (reasoning
layer) from a \emph{DevBox} (execution VM with shell, IDE, and browser).
A persistent Knowledge store provides cross-session memory. Devin is the one
system surveyed here with no accompanying paper: the only primary source is a
launch blog post, which reports 13.86\% resolved on SWE-bench. That is an order
of magnitude above the 1.96\% retrieval-augmented baseline, and it is the result
that launched the agentic turn, but it was measured on a random 25\% subset
rather than the full test set, with no released trajectories or harness. The
claim has never been independently reproduced. The one longitudinal external
assessment, a month-long trial by Answer.AI, found Devin completed 3~of 20~tasks,
with 14~outright failures~\cite{answerai2025devin}. We include Devin because of its
architectural influence and its role in the benchmark's history, not as a
comparable data point; its row in Table~\ref{tab:e2e-systems} should be read
as a vendor claim rather than a measurement.

Taken together, these five systems expose a recurring trade-off between
agentic flexibility and pipeline efficiency. Agentless shows that a fixed
localise--repair--validate pipeline can outperform interactive agents on
SWE-bench Lite at lower cost, challenging the assumption that autonomy is
necessary for strong results. This claim needs care, because the rows of
Table~\ref{tab:e2e-systems} do not share a backbone: Agentless runs on GPT-4o
and SWE-agent's published figure on GPT-4~Turbo, and \S\ref{sec:e2e-eval} shows
that swapping the backbone alone can move a resolve rate by ten points. The
claim survives only because Agentless's authors also report the matched
comparison, running the agentic baselines on the same GPT-4o backbone, and the
gap persists there~\cite{xia2024agentless}. It is that matched comparison, not
the cross-backbone table, that carries the argument. Among agentic designs, interface
and context strategy matter as much as the backbone LLM: SWE-agent's ACI
ablations, OpenHands's CodeAct action space, and AutoCodeRover's AST search
APIs each demonstrate that how an agent reads and edits code can shift resolve
rates by several percentage points at comparable model capacity.

\subsection{Full-Pipeline Evaluation}
\label{sec:e2e-eval}

The \emph{Evaluation} branch of Figure~\ref{fig:survey-taxonomy} summarises the
benchmark splits, scoring protocols, and confounds that shape current
performance reports.

\paragraph{SWE-bench.}
The primary evaluation instrument for end-to-end agents is
SWE-bench~\cite{jimenez2024swebench}: 2{,}294~merged pull requests scraped from
12~Python repositories, retaining only PRs that both resolve a linked GitHub
issue and modify the project's test suite. The two filters do different work.
The test-suite filter supplies the grading oracle; the issue-linkage filter
supplies the task's \emph{intent}, and it is the reason an agent has any
natural-language input at all. An agent receives the repository snapshot from
\emph{before} the merge together with the
issue text; its patch is applied and scored against \emph{fail-to-pass} tests
(which must flip) and \emph{pass-to-pass} tests (which must not regress). The
headline metric throughout this report is the \emph{resolve rate}: the
percentage of instances for which both test sets pass on the first submitted
patch (pass@1).

\paragraph{Terminology and comparability.}
The name ``SWE-bench'' denotes a family of task sets rather than a single
benchmark, and its splits differ in construction as much as in size.
Table~\ref{tab:swebench-splits} summarises the vocabulary adopted throughout this
report. Two distinctions warrant emphasis. First, Verified is the most carefully
curated split rather than the most demanding one: human screening removed
underspecified issues and unfair tests, so resolve rates on Verified are
systematically higher than on Full or Lite. Scores are consequently not
comparable across splits, and the contemporary frontier is measured more
informatively on Pro, which is both harder and constructed to resist
memorisation. Second, the labels \emph{easy}, \emph{medium}, and \emph{hard} do
not designate splits. They are annotator difficulty tiers defined within
Verified according to the time a human engineer was estimated to require
($<$15~min, 15~min--1~h, $>$1~h)~\cite{ganhotra2025difficulty}.

\begin{table}[t]
  \centering
  \caption{The SWE-bench family. Lite and Verified are subsets of Full and share
    its 12~Python repositories; Multilingual and Pro are independently
    constructed task sets that reuse the harness. Resolve rates are
    \emph{not} comparable across rows.}
  \label{tab:swebench-splits}
  \small
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{@{}llp{4.5cm}@{}}
    \toprule
    \textbf{Split} & \textbf{Size} & \textbf{Selection criterion} \\
    \midrule
    Full & 2{,}294 & All scraped PRs that touch tests \\
    Lite & 300 & Single-file patches, self-contained issues; some issue texts
      leak the solution~\cite{xia2024agentless} \\
    Verified & 500 & Human-screened for issue clarity and test
      fairness~\cite{openai2024verified} \\
    Multilingual & 300 & Same pipeline, 9~non-Python
      languages~\cite{swebench2025multilingual} \\
    Pro & 1{,}865 & Long-horizon multi-file changes; copyleft and held-out
      proprietary repos~\cite{deng2025swebenchpro} \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Task complexity profile.}
The retained tasks sit at the small end of real-world software
engineering~\cite{ganhotra2025difficulty,epoch2025swebench}. Annotators judged
roughly nine in ten Verified instances fixable in under an hour, and even the
hard tier averages well under a hundred changed lines. Instances are
overwhelmingly single-file bug fixes rather than feature work or refactorings,
and heavily concentrated in a few repositories, with Django alone contributing
close to half. SWE-bench therefore measures localised repair on mature Python
codebases rather than software engineering in the broad sense, a caveat that
shapes how its leaderboard should be read.

\paragraph{Leaderboard trajectory.}
At launch, retrieval-augmented prompting of a frontier model scored barely above
zero~\cite{jimenez2024swebench}: the models could write plausible patches but had
no way to find the right file or check their own work. The first agentic
scaffolds (SWE-agent, AutoCodeRover, Devin) moved the frontier by an order of
magnitude within months without any change to the underlying model, purely by
supplying tools, a repository to navigate, and tests to run against. The 2024
scaffolding race that followed exposed how much of that gain came from
scaffolding rather than reasoning capacity, since a deliberately simple
localise-then-repair pipeline such as Agentless stayed competitive with far more
elaborate agentic loops.

A headline resolve rate hides where that progress actually landed.
Figure~\ref{fig:swebench-trajectory} therefore splits Verified by the difficulty
its human annotators assigned, tracking nine leaderboard systems from
April~2024 to March~2025.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      width=\columnwidth,
      height=6.6cm,
      xlabel={Leaderboard submission date},
      ylabel={\% Resolved},
      ymin=0, ymax=104,
      ytick={0,20,40,60,80,100},
      xmin=2024.16, xmax=2025.30,
      xtick={2024.25, 2024.5, 2024.75, 2025.0, 2025.25},
      xticklabels={Apr'24, Jul'24, Oct'24, Jan'25, Apr'25},
      xticklabel style={font=\scriptsize},
      yticklabel style={font=\scriptsize},
      ylabel style={font=\small},
      xlabel style={font=\small},
      grid=major,
      grid style={dashed, gray!30},
      legend style={font=\tiny, at={(0.5,1.02)}, anchor=south,
                    legend columns=2, draw=none, fill=none,
                    /tikz/every even column/.append style={column sep=5pt},
                    row sep=1pt},
      every axis plot/.append style={thick, mark size=1.5pt},
    ]
    % Resolve rates: Ganhotra (2025), single source; tiers n=194/261/45 of 500.
    % x = submission date from the leaderboard's dated result directories
    % (SWE-bench/experiments, evaluation/verified/<YYYYMMDD>_<system>).
    %   2024.251 SWE-agent + Claude 3 Opus     (20240402)
    %   2024.467 SWE-agent + Claude 3.5 Sonnet (20240620)
    %   2024.571 SWE-agent + GPT-4o            (20240728)
    %   2024.918 Agentless-1.5 + Sonnet 3.5    (20241202)
    %   2024.918 Amazon Q Developer Agent      (20241202)
    %   2024.975 Emergent E1                   (20241223)
    %   2025.044 W&B Programmer O1 crosscheck5 (20250117)
    %   2025.099 AgentScope                    (20250206)
    %   2025.203 Augment Agent v0              (20250316)
    \addplot[green!50!black, mark=*] coordinates {
      (2024.251,27.3) (2024.467,47.9) (2024.571,36.6) (2024.918,70.6)
      (2024.918,72.2) (2024.975,74.7) (2025.044,77.3) (2025.099,81.4)
      (2025.203,80.4)
    };
    \addlegendentry{Easy, $<$15\,min ($n{=}194$)}

    \addplot[black, mark=square*] coordinates {
      (2024.251,18.2) (2024.467,33.6) (2024.571,23.2) (2024.918,50.8)
      (2024.918,55.0) (2024.975,57.2) (2025.044,64.6) (2025.099,63.4)
      (2025.203,65.4)
    };
    \addlegendentry{Overall ($n{=}500$)}

    \addplot[blue!65!black, mark=triangle*] coordinates {
      (2024.251,10.0) (2024.467,28.0) (2024.571,16.9) (2024.918,42.5)
      (2024.918,49.4) (2024.975,50.6) (2025.044,62.1) (2025.099,56.7)
      (2025.203,62.1)
    };
    \addlegendentry{Medium, 15\,min--1\,h ($n{=}261$)}

    \addplot[red!70!black, mark=diamond*] coordinates {
      (2024.251,0.0) (2024.467,4.4) (2024.571,2.2) (2024.918,13.3)
      (2024.918,13.3) (2024.975,20.0) (2025.044,24.4) (2025.099,24.4)
      (2025.203,20.0)
    };
    \addlegendentry{Hard, $\geq$1\,h ($n{=}45$)}

    % Union over all nine systems (same source): easy 95.36, hard 42.22.
    \draw[dashed, gray!65] (axis cs:2024.16,95.36) -- (axis cs:2025.30,95.36);
    \node[font=\tiny, text=gray!40!black, anchor=east, fill=white, inner sep=1pt]
      at (axis cs:2025.28,90.3) {union of all nine: easy 95.4\%};
    \draw[dashed, gray!65] (axis cs:2024.16,42.22) -- (axis cs:2025.30,42.22);
    \node[font=\tiny, text=gray!40!black, anchor=east, fill=white, inner sep=1pt]
      at (axis cs:2025.28,37.0) {union of all nine: hard 42.2\%};

    % Annotate the backbone dip (same scaffold, weaker model).
    \node[font=\tiny, text=gray!30!black, anchor=north, align=center,
          inner sep=1pt] at (axis cs:2024.60,16.0) {SWE-agent\\+ GPT-4o};
    \end{axis}
  \end{tikzpicture}
  \caption{SWE-bench Verified resolve rate over time, overall and split by
    annotator-estimated difficulty, for the nine systems whose per-tier results
    have been published. Resolve rates and difficulty tiers come from a single
    re-analysis of the leaderboard's predictions~\cite{ganhotra2025difficulty},
    using the human annotations shipped with
    Verified~\cite{openai2024verified}, so the four series are directly
    comparable. Each point is placed at that submission's date in the
    leaderboard's result archive~\cite{swebench_leaderboard}. The curves connect
    the nine systems in submission order; they are \emph{not} a running
    best-in-class frontier, which is why they dip in July~2024, when SWE-agent
    was resubmitted with a weaker backbone than the Claude~3.5~Sonnet run of the
    previous month. Dashed lines mark the \emph{union} over all nine systems:
    the instances solved by at least one of them. The overall series is the
    tier-weighted average of the other three, with one exception: for
    SWE-agent~+~Claude~3~Opus the source's headline 18.2\% (91/500) exceeds the
    sum of its own per-tier counts (79/500).}
  \label{fig:swebench-trajectory}
\end{figure}

Three things stand out, and none of them is visible in the overall curve. First,
the aggregate score is dominated by composition. Easy and medium tasks are 91\%
of Verified, so the headline number mostly reports how well an agent handles
work a developer would finish within an hour. Second, the tiers improved at very
different rates. Across the nine systems the easy tier rose from 27.3\% to
80.4\%, a factor of 2.9, while the medium tier rose from 10.0\% to 62.1\%, a
factor of 6.2. The easy tier is now near its ceiling, so almost all recent
movement in the overall score is coming from the medium tier. Third, the hard
tier behaves differently again: it climbed from 0\% to about 20\% by the end of
2024 and then stopped, with the last three systems scoring 24.4\%, 24.4\% and
20.0\%. A system can gain ten points overall while solving no additional hard
task, which is what makes the overall number a poor proxy for progress on the
tasks that matter most.

The union sharpens this. Pooled across all nine systems, some agent solves 95.4\%
of easy and 84.3\% of medium instances but only 42.2\% of hard ones, and just
9 of the 25 hard multi-file issues~\cite{ganhotra2025difficulty}. The remaining
easy and medium headroom is therefore largely an ensembling problem: different
scaffolds already solve different instances, which is why the inference-time
scaling discussed in \S\ref{sec:discussion} pays off. The hard tier is not an
ensembling problem. More than half of it defeats every system at once, which
points to a capability ceiling rather than a sampling one.

Two caveats bound how far the figure can be read. First, it plots nine
independent submissions in date order, not a running best-in-class frontier, and
it stops in March~2025 because the re-analysis behind it does. The July~2024 dip
shows why the distinction matters: the same SWE-agent scaffold resolves 33.6\%
with a Claude~3.5~Sonnet backbone in June and 23.2\% with GPT-4o six weeks later.
Scaffolding is not the only thing that moves these curves. Lite and Full,
meanwhile, receive few submissions at all now that teams have migrated to
Verified~\cite{dissecting2025}, and the best rates ever recorded on them, 60.3\%
and 52.6\%~\cite{swebench_leaderboard}, reflect that thinning field as much as
the difficulty of the splits themselves.
Second, contamination sits on top of everything: half the issues predate 2020, in
repositories that are staple training data~\cite{epoch2025swebench}. Frontier
models score tens of points lower on Pro, whose problems are novel and built to
resist memorisation, than on Verified;
\S\ref{sec:eval-landscape} quantifies that gap and asks how much of it can
honestly be attributed to memorisation.

\paragraph{Beyond SWE-bench.}
Pro and Multilingual (Table~\ref{tab:swebench-splits}) answer the contamination
and language critiques from within the SWE-bench format. A separate line of work
leaves the format altogether and scores agents by the compute they spend rather
than by resolve rate alone~\cite{sweeffi2025}. \S\ref{sec:eval-landscape} treats
the evaluation landscape in full.

\section{Planning and Search Strategies for Code Synthesis}
\label{sec:planning}
Planning is the stage in which an agent turns a vague intent into an
operational hypothesis about \emph{what} code should change and \emph{how} the
change can be produced. In the vibe coding pipeline, this stage sits between
intent capture and code generation, but in practice it is rarely a single
up-front step. Agents plan while reading issue reports, searching repositories,
choosing files, decomposing edits, sampling candidate patches, and revising
their strategy after failed tests. Thus, planning is best understood as a
control layer over generation rather than as a separate document written before
generation begins. While \S\ref{sec:e2e} evaluates end-to-end agents primarily
on SWE-bench, the planning and search mechanisms below are studied across
function-level benchmarks (HumanEval, MBPP), competitive programming
(CodeContests), and repository repair tasks alike. This section instantiates
the \emph{Planning \& Search} branch of Figure~\ref{fig:survey-taxonomy}.

This section surveys four aspects of that control layer:
\begin{enumerate}
  \item Intent capture: how the natural-language task enters the pipeline,
    and what happens when it is underspecified;
  \item Plan representations: determine what intermediate state the
    agent exposes to itself: natural-language rationales, hierarchical task
    lists, program sketches, or repository-localisation hypotheses;
  \item Search strategies: determine how many possible solutions are
    explored and how candidates are ranked;
  \item The plan-generate interface: determines whether planning happens
    once, repeatedly, or only when execution feedback invalidates the current
    hypothesis.
\end{enumerate}

\begin{table}[t]
  \centering
  \caption{Planning and search mechanisms in LLM-based code synthesis.}
  \label{tab:planning-search}
  \scriptsize
  \begin{tabular}{@{}p{0.22\columnwidth}p{0.31\columnwidth}p{0.34\columnwidth}@{}}
    \toprule
    \textbf{Mechanism} & \textbf{Planning object} & \textbf{Role in code tasks} \\
    \midrule
    Chain / tree / graph of thoughts
      & Intermediate reasoning steps
      & Explore alternative designs, algorithms, or debugging hypotheses before emitting code \\
    Hierarchical decomposition
      & Subtasks, files, functions, edit locations
      & Reduce repository-scale tasks into bounded local edits \\
    Candidate sampling
      & Multiple complete or partial programs
      & Trade inference cost for higher pass@k and diversity \\
    Execution-guided selection
      & Tests, generated tests, compiler output
      & Rank or prune candidates using external evidence rather than model confidence alone \\
    ReAct-style replanning
      & Updated belief state after tool observations
      & Interleave search, reading, editing, and feedback in a single loop \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Intent Capture: The Open-Loop Stage}
\label{sec:intent}

Intent enters the pipeline as a GitHub issue, a bug report, or a sentence typed
into a chat box. It is the only stage of Figure~\ref{fig:vibe-pipeline} that no
surveyed system instruments, and the asymmetry is worth dwelling on, because it
is not obviously the right design.

\paragraph{Intent quality dominates outcomes.}
The clearest evidence comes from the benchmarks themselves. SWE-bench Verified
was built by having human engineers discard instances whose issue text was
\emph{underspecified}, that is, instances where the intent did not determine the
patch~\cite{openai2024verified}. Screening the intent, while changing neither
the repositories, the agents, nor the grading harness, raises resolve rates by
tens of points (\S\ref{sec:e2e-eval}). The same lever appears with the opposite
sign in Lite, where some issue descriptions leak the fix outright, so that the
intent contains its own solution~\cite{xia2024agentless}. Between these two
poles lies the entire difficulty range that Figure~\ref{fig:swebench-trajectory}
stratifies. An agent's score is, to a degree that is rarely acknowledged, a
score on the issue-writing of the repository it was sampled from.

\paragraph{No agent closes the loop here.}
Every other stage of the pipeline regulates on an external signal. Planning is
checked against repository structure (does \texttt{search\_class} find the
symbol?); generation against a parser and a linter; execution against a test
runner; refinement against the resulting failures. Intent capture has no such
signal. None of the five systems in \S\ref{sec:e2e-systems} asks a clarifying
question when the issue text is ambiguous; each instead commits to a reading of
it and proceeds. OpenHands and Devin admit human interjection
(\S\ref{sec:convergence}), but the initiative is the human's: the agent does not
detect underspecification and request disambiguation. In control terms the
stage is \emph{open-loop}. Where the pipeline everywhere else substitutes an
external oracle for the model's own confidence, at the intent stage the model's
reading of the issue simply \emph{is} the specification, unchallenged.

\paragraph{Why the omission persists.}
The benchmark makes the omission invisible. A dataset of issues that were, by
construction, resolved by a real merged PR is a dataset in which the intent was
adequate; and Verified deliberately removed the instances where it was not. An
agent is therefore never rewarded for asking, and never penalised for guessing
wrong about a genuinely ambiguous request. The cost surfaces only outside the
benchmark, as the specification-ambiguity challenge of \S\ref{sec:open}. We
regard this as the clearest structural gap the pipeline framing exposes: four
stages have been given feedback, and the fifth has been given a leaderboard that
hides the fact that it has none.

\subsection{Plan Representations and Decomposition}

\begin{figure}[t]
  \centering
  \resizebox{\columnwidth}{!}{%
  \begin{tikzpicture}[
    node distance=0.8cm and 1.2cm,
    thought/.style={draw, circle, fill=blue!10, minimum size=0.6cm, font=\tiny},
    arr/.style={-{Stealth[length=1.2mm]}, semithick},
    backarr/.style={-{Stealth[length=1.2mm]}, semithick, dashed},
  ]
    % Chain of Thought
    \node[thought] (c1) {A};
    \node[thought, right=of c1] (c2) {B};
    \node[thought, right=of c2] (c3) {C};
    \node[thought, right=of c3] (c4) {D};
    \draw[arr] (c1) -- (c2);
    \draw[arr] (c2) -- (c3);
    \draw[arr] (c3) -- (c4);
    \node[above=0.15cm of c2, font=\scriptsize\bfseries] {Chain of Thought};
    
    % Tree of Thoughts
    \node[thought, below=2cm of c1] (t1) {A};
    \node[thought, below right=0.5cm and 0.4cm of t1] (t2a) {B$_1$};
    \node[thought, below left=0.5cm and 0.4cm of t1] (t2b) {B$_2$};
    \node[thought, below right=0.5cm and 0.2cm of t2a] (t3a) {C$_1$};
    \node[thought, below left=0.5cm and 0.2cm of t2a] (t3b) {C$_2$};
    \node[thought, below=0.7cm of t2b] (t3c) {C$_3$};
    \draw[arr] (t1) -- (t2a);
    \draw[arr] (t1) -- (t2b);
    \draw[arr] (t2a) -- (t3a);
    \draw[arr] (t2a) -- (t3b);
    \draw[arr] (t2b) -- (t3c);
    \draw[backarr, red!60] (t3b) to[bend right=45] (t2a);
    \node[above=0.15cm of t1, font=\scriptsize\bfseries] {Tree of Thoughts};
    \node[below=0.05cm of t3b, font=\tiny, text=red!70] {backtrack};
    
    % Graph of Thoughts
    \node[thought, below=2cm of c3] (g1) {A};
    \node[thought, below right=0.5cm and 0.4cm of g1] (g2a) {B$_1$};
    \node[thought, below left=0.5cm and 0.4cm of g1] (g2b) {B$_2$};
    \node[thought, below=1.2cm of g1] (g3) {C};
    \node[thought, below=0.7cm of g3] (g4) {D};
    \draw[arr] (g1) -- (g2a);
    \draw[arr] (g1) -- (g2b);
    \draw[arr, blue!70] (g2a) -- (g3);
    \draw[arr, blue!70] (g2b) -- (g3);
    \draw[arr] (g3) to[bend right=40]
      node[midway, right, font=\tiny, inner sep=1pt] {revisit} (g2a);
    \draw[arr] (g3) -- (g4);
    \node[above=0.15cm of g1, font=\scriptsize\bfseries] {Graph of Thoughts};
    \node[left=0.05cm of g3, font=\tiny, text=blue!70] {merge};
  \end{tikzpicture}%
  }
  \caption{Plan representation structures. Chain of Thought commits to a single
    linear path. Tree of Thoughts explores alternatives and backtracks. Graph of
    Thoughts merges branches and revisits earlier states.}
  \label{fig:plan-shapes}
\end{figure}

Plan representations determine how an agent organises reasoning before generating
code. The simplest is chain of thought~\cite{wei2022chain}: a linear sequence
that commits early to one trajectory. If the first design choice is wrong,
subsequent steps elaborate the mistake. Tree of Thoughts~\cite{yao2023tree} and
Graph of Thoughts~\cite{besta2024graph} relax this by allowing exploration,
backtracking, and merging across alternative paths (Figure~\ref{fig:plan-shapes}).
The mechanism in Tree of Thoughts is worth naming precisely, since the figure's
``backtrack'' arrow understates it: the tree is expanded under a classical
search policy (breadth- or depth-first) in which an LLM \emph{value function}
scores each partial state, so branches are pruned by lookahead rather than
abandoned only once they fail. The evaluator, not the branching, is what makes
the structure useful, and it is also what a repository-scale agent lacks, since
no cheap value function scores a half-finished patch.
For code synthesis, this means agents can consider multiple decomposition
strategies (``change the parser'', ``change the caller'', or ``adjust the
compatibility layer'') as explicit search states rather than implicit in a
single sample.

Repository-level agents usually add a more concrete representation
(recall \S\ref{sec:e2e-systems}): \emph{localisation plans}. Agentless
decomposes issue resolution into files,
classes/functions, and edit locations before generating a
patch~\cite{xia2024agentless}. AutoCodeRover uses AST-backed APIs such as
\texttt{search\_class} and \texttt{search\_method\_in\_class} to plan over
program structure rather than raw text~\cite{zhang2024autocoderover}. These
grounded representations name concrete program elements that can be validated
against the repository (Figure~\ref{fig:localisation-pipeline}), turning a
broad intent into a bounded search problem.

\begin{figure}[t]
  \centering
  \resizebox{0.9\columnwidth}{!}{%
  \begin{tikzpicture}[
    node distance=0.6cm,
    box/.style={draw, rounded corners=2pt, fill=green!8, font=\scriptsize,
                align=center, minimum height=0.55cm, minimum width=1.8cm},
    arr/.style={-{Stealth[length=1.2mm]}, semithick},
  ]
    \node[box] (issue) {Issue\\text};
    \node[box, right=of issue] (files) {Candidate\\files};
    \node[box, right=of files] (funcs) {Classes/\\functions};
    \node[box, right=of funcs] (locs) {Edit\\locations};
    \node[box, right=of locs] (patch) {Patch};
    
    \draw[arr] (issue) -- (files);
    \draw[arr] (files) -- (funcs);
    \draw[arr] (funcs) -- (locs);
    \draw[arr] (locs) -- (patch);
  \end{tikzpicture}%
  }
  \caption{Hierarchical localisation pipeline. Agents narrow scope
    iteratively from repository-level to line-level edits.}
  \label{fig:localisation-pipeline}
\end{figure}

Plans may also be expressed as \emph{specification sketches}: partial
interfaces, pseudocode, assertions, or examples that constrain the generated
implementation. Useful plans are grounded in observable artefacts (issue text,
existing code, API conventions, and failing tests) rather than invented
requirements.

\subsection{Search and Exploration Strategies}

Once a representation exists, an agent must decide how broadly to search.
Single-sample generation is fast, but it treats the model's first completion as
the solution. Search-based systems instead spend additional inference or
execution budget to explore alternatives. The simplest form is self-consistency:
sample multiple reasoning paths and select the most common or most consistent
answer~\cite{wang2023selfconsistency}. Transposing this to code forces the idea
to mutate, and the reason is instructive. Self-consistency marginalises over
reasoning paths that terminate in a \emph{single canonical answer}, so votes can
be counted by string equality. Programs have no canonical form: two correct
patches may share no token, and two textually near-identical patches may differ
semantically. The vote must therefore be taken in a different space. Systems
recover it either \emph{behaviourally}, by clustering candidates on their
outputs over a set of inputs (AlphaCode, CodeT), or \emph{syntactically up to
normalisation}, by comparing candidates after AST canonicalisation
(Agentless~\cite{xia2024agentless}). Sample-and-rank is thus self-consistency
with an equivalence relation borrowed from program semantics rather than from
string matching, and the quality of that relation bounds what the vote can buy.

\begin{figure}[t]
  \centering
  \resizebox{\columnwidth}{!}{%
  \begin{tikzpicture}[
    node distance=0.8cm,
    box/.style={draw, rounded corners=2pt, fill=blue!8, font=\scriptsize,
                align=center, minimum height=0.6cm, minimum width=2.2cm},
    arr/.style={-{Stealth[length=1.4mm]}, semithick},
  ]
    \node[box] (gen) {Generate\\$N$ candidates};
    \node[box, right=of gen] (exec) {Execute \&\\test};
    \node[box, right=of exec] (cluster) {Cluster \&\\rank};
    \node[box, right=of cluster] (select) {Select\\patch};
    
    \draw[arr] (gen) -- (exec);
    \draw[arr] (exec) -- (cluster);
    \draw[arr] (cluster) -- (select);
    
    \node[below=0.25cm of gen, font=\tiny, text width=2.2cm, align=center]
      {LLM samples\\or mutations};
    \node[below=0.25cm of exec, font=\tiny, text width=2.2cm, align=center]
      {unit tests,\\generated tests};
    \node[below=0.25cm of cluster, font=\tiny, text width=2.2cm, align=center]
      {AST similarity,\\pass/fail groups};
    \node[below=0.25cm of select, font=\tiny, text width=2.2cm, align=center]
      {top-1 or\\small ensemble};
  \end{tikzpicture}%
  }
  \caption{Candidate-selection pipeline. Systems like AlphaCode and CodeT
    generate many candidates, execute them against tests (real or generated),
    cluster by behaviour, and select a small high-quality set.}
  \label{fig:candidate-selection}
\end{figure}

AlphaCode~\cite{li2022alphacode} and CodeT~\cite{chen2022codet} exemplify this
pattern (Figure~\ref{fig:candidate-selection}). AlphaCode generated millions of
candidates for competitive programming, filtered and clustered by execution
results, then submitted a small diverse set. CodeT generates both programs and
tests, selecting code consistent with generated test cases. The key insight is
that LLMs often contain the right solution somewhere in the sample distribution
even when top-ranked output is wrong. The limitation is self-referential
objectives: if generated tests are weak, selection optimises for the model's
own misunderstanding. Execution-guided search is strongest when feedback
includes external oracles (existing unit tests, compiler errors, type checkers,
or repository-specific regression suites).

Search also appears in repository navigation, as in the systems surveyed in
\S\ref{sec:e2e}. Agentless samples multiple locations and candidate patches,
then normalises patches by AST and uses validation to pick among
them~\cite{xia2024agentless}. AutoCodeRover expands
context iteratively through program-structure queries, optionally using
Spectrum-Based Fault Localisation (SBFL) to prioritise suspicious code when
tests are available~\cite{zhang2024autocoderover}. In both cases, search is
not over complete programs alone; it is also over \emph{where to look} and
\emph{what evidence to include}. This matters because code generation quality
is bounded by context quality. A correct patch is unlikely if the agent never
places the relevant function, invariant, or failing trace in context.

The cost of search is the main trade-off. More samples increase diversity and
pass@k, but they also increase latency, token cost, and validation cost.
Search can also overfit to narrow tests: a candidate that passes the
available suite may still violate unstated requirements. Strong systems
therefore combine multiple pruning signals: syntactic validity, static
analysis, failing-test repair, pass-to-pass regression preservation, and
semantic similarity to the intended behaviour. Search is most useful when the
ranking signal is at least as reliable as the generator is diverse.

\subsection{Plan--Generate Interplay}

Planning and generation can be coupled in three broad ways
(Table~\ref{tab:plan-generate}). \emph{Plan-then-generate} is easy to inspect
and aligns with human practice, but brittle if the initial plan is
wrong~\cite{xia2024agentless}. \emph{Interleaved planning} adapts to new
information but distributes the plan across many turns, making it harder to
audit~\cite{yao2023react}. \emph{Generate-then-refine} treats the first output
as a hypothesis and revises via self-critique or execution
feedback~\cite{madaan2023selfrefine}, connecting planning to the refinement
loop in \S\ref{sec:refinement}.

\begin{table}[t]
  \centering
  \caption{Plan--generate coupling patterns.}
  \label{tab:plan-generate}
  \small
  \begin{tabular}{@{}p{0.23\columnwidth}p{0.35\columnwidth}p{0.28\columnwidth}@{}}
    \toprule
    \textbf{Pattern} & \textbf{Mechanism} & \textbf{Trade-off} \\
    \midrule
    Plan-then-generate
      & Plan once, then generate code conditioned on it
      & Easy to inspect; brittle if plan is wrong \\
    Interleaved planning
      & Alternate plan updates and tool use in observe--think--act loop
      & Adaptive; plan distributed across turns \\
    Generate-then-refine
      & Generate, critique via feedback, revise iteratively
      & Self-correcting; may reopen planning after execution \\
    \bottomrule
  \end{tabular}
\end{table}

The central design question is not whether agents should plan, but how plans
remain \emph{grounded}. Grounding mechanisms include retrieval over repository
structure, AST-level symbol search, execution traces, failing tests, and
explicit constraints extracted from the issue report. Effective systems keep
plans close to these artefacts and treat generation as one move in a larger
search process, reframing code synthesis from ``write the answer'' to
``maintain and test a hypothesis about the required program change''.

This distinction resolves what would otherwise be an embarrassment for the
control-layer thesis: Agentless, the system that plans \emph{least}, is
competitive with those that plan most (\S\ref{sec:e2e-systems}). If the claim
were that deliberation helps, Agentless would refute it. The claim is instead
that \emph{grounding} helps, and deliberation is merely one way to obtain it,
often an expensive one. Agentless obtains grounding structurally, by hard-coding
the hierarchical narrowing that an agentic loop must rediscover through
exploration, and it therefore confirms the thesis rather than embarrassing it.
The corollary is that finding the right location is harder than editing it once
found: AutoCodeRover's SBFL variant changes nothing about repair, only about
localisation, and buys three points of resolve rate together with seven
instances no other configuration reaches~\cite{zhang2024autocoderover}. Planning
in this setting is not reasoning about \emph{how} to write code. It is reasoning
about \emph{where} the code is, and about what evidence would confirm that guess.

\section{Execution Feedback and Iterative Refinement}
\label{sec:refinement}

Building on the plan--generate patterns surveyed in \S\ref{sec:planning}, this
section examines how execution closes the loop. Code generation rarely succeeds
in one shot. Even human programmers iterate: write, compile, test, debug,
revise. Agentic systems execute generated code, observe runtime
outcomes (test failures, compiler errors, exceptions), and revise their output
accordingly. While \S\ref{sec:e2e} scores agents on SWE-bench, refinement
mechanisms are often evaluated on function-level suites (HumanEval, MBPP),
text-to-SQL (Spider), and APR benchmarks (Defects4J). This section instantiates
the \emph{Execution Feedback} branch of Figure~\ref{fig:survey-taxonomy} and
examines the feedback signals agents consume (\S\ref{sec:feedback-signals}),
the mechanisms by which they refine code (\S\ref{sec:refinement-mechanisms}),
the evidence on whether refinement actually works
(\S\ref{sec:refinement-limits}), the techniques for localising and repairing
faults (\S\ref{sec:fault-localization}), and the criteria for deciding when to
stop iterating (\S\ref{sec:convergence}).

One distinction organises the whole section, and we state it up front rather
than deriving it late: \emph{what closes the loop}. A refinement mechanism is
only a control loop if some signal outside the model's own judgement can
contradict the model. Where that signal is an external oracle (a test runner, a
compiler, a type checker) refinement reliably improves code. Where the signal is
the model's own critique of its own output, the loop is closed onto itself and
the apparent control is illusory. Table~\ref{tab:loop-closure} sorts the
mechanisms surveyed below on exactly this axis, and \S\ref{sec:refinement-limits}
supplies the evidence that the axis is the one that matters.

\subsection{Feedback Signals and Environments}
\label{sec:feedback-signals}

Agents require feedback to improve. The type, source, and presentation of
feedback critically affect refinement efficacy.

\paragraph{Types of feedback.}
The most common feedback is \emph{test execution results}: pass/fail verdicts,
failing test cases, assertion messages, and expected-versus-actual output diffs.
When tests fail, agents receive concrete evidence of incorrectness. A richer
signal is \emph{compiler or interpreter errors}: syntax errors, type
mismatches, undefined-variable messages, and stack traces. These pinpoint the
location and nature of faults without requiring test design. Static analysis
tools provide another layer: linters flag style violations, type checkers
identify inconsistencies, and data-flow analysers detect null dereferences or
race conditions before execution. Some systems incorporate \emph{human feedback}
in the loop: developers approve or reject patches, provide natural-language
explanations of failures, or annotate bug reports. Finally, agents can
generate \emph{self-critique}: an LLM explains the generated code in natural
language, identifies logical flaws, or predicts potential failures without
executing the code (``rubber duck debugging'')~\cite{chen2024selfdebug}.

\paragraph{Feedback sources: external versus internal.}
Feedback may come from \emph{external} oracles (pre-written test suites,
continuous-integration pipelines, runtime environments) or be
\emph{internally simulated} by the agent itself. The distinction is not a
taxonomy convenience; it is the load-bearing variable of this section. External
feedback is authoritative because it can \emph{surprise} the model: a test that
fails contradicts the model's belief regardless of how confident that belief
was. It requires infrastructure and may be slow or expensive. Internal feedback
is fast and requires no setup, but it is generated by the same distribution that
produced the error, so it inherits the error's blind spots: the agent may
confidently believe its code is correct, or fabricate an error message for a
line that is fine. An agent that could reliably tell its wrong code from its
right code by inspection would have produced the right code in the first place.

Table~\ref{tab:loop-closure} sorts the mechanisms of
\S\ref{sec:refinement-mechanisms}--\S\ref{sec:fault-localization} by what closes
their loop. Hybrid approaches, which use external execution for validation and
internal self-critique for \emph{diagnosis}, occupy the productive middle: the
oracle decides \emph{whether} the code is wrong, the model proposes \emph{why}.
\S\ref{sec:refinement-limits} argues that this division of labour, and not
self-critique as such, is what makes refinement work.

\begin{table}[t]
  \centering
  \caption{Refinement mechanisms sorted by what closes the loop. Only an
    external oracle can contradict the model; internal signals can only
    reorganise what the model already believes.}
  \label{tab:loop-closure}
  \scriptsize
  \setlength{\tabcolsep}{3pt}
  \begin{tabular}{@{}p{0.24\columnwidth}p{0.16\columnwidth}p{0.24\columnwidth}p{0.26\columnwidth}@{}}
    \toprule
    \textbf{Mechanism} & \textbf{Loop closed by} & \textbf{Signal}
      & \textbf{Characteristic failure} \\
    \midrule
    Self-Refine~\cite{madaan2023selfrefine}
      & Internal & Model critiques own output
      & Plateaus; no signal can contradict the generator \\
    Self-debugging, no tests~\cite{chen2024selfdebug}
      & Internal & Self-generated line-by-line explanation
      & Explains the bug as intended behaviour \\
    Self-debugging, with tests~\cite{chen2024selfdebug}
      & Hybrid & Test verdict $+$ model diagnosis
      & Bounded by test coverage \\
    Reflexion~\cite{shinn2023reflexion}
      & Hybrid & External reward/verdict, verbalised into memory
      & Reward must exist and be dense enough \\
    SelfEvolve~\cite{jiang2023selfevolve}
      & External & Interpreter errors, traces
      & Silent (non-crashing) semantic bugs \\
    D4C~\cite{xu2024d4c}
      & External & Failing tests $+$ full function
      & Plausible-but-incorrect patches \\
    Agentless validate~\cite{xia2024agentless}
      & External & Regression $+$ generated tests
      & Generated tests inherit the model's misreading \\
    Best-of-$N$ $+$ critic~\cite{openhands2025verified}
      & Hybrid & Executed rollouts ranked by learned critic
      & Cost scales with $N$ \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Sandboxed execution environments.}
To safely run untrusted generated code, systems employ sandboxing. Common
techniques include \emph{containerisation} (Docker), \emph{virtual machines},
\emph{language-level isolation} (restricted Python interpreters), and
\emph{timeout and resource limits}. SWE-agent~\cite{yang2024sweagent} and
AutoCodeRover~\cite{zhang2024autocoderover} execute each task inside a
per-instance Docker container, which is what makes SWE-bench's
environment-dependent test suites reproducible at all; OpenHands additionally
runs a browser and an IPython server in the same
sandbox~\cite{wang2024openhands}. Sandboxing serves two purposes that are easily
conflated: \emph{reproducibility}, ensuring the harness scores the patch and not
the environment, and \emph{containment}, preventing generated code from damaging
the host. Most research systems are engineered for the former; the latter
matters once agents run against real infrastructure
(\S\ref{sec:open}). Both introduce latency and restrict access to the databases
and network services that production code needs, so balancing safety and realism
remains an open design challenge.

\subsection{Iterative Refinement Mechanisms}
\label{sec:refinement-mechanisms}

Once feedback is available, agents must incorporate it to produce better code.
Refinement strategies vary in how they represent feedback, how they prompt the
model to revise, and whether they maintain memory across iterations.

\paragraph{Reflexion: verbal reinforcement learning.}
Reflexion~\cite{shinn2023reflexion}, from Northeastern, MIT, and Princeton,
treats refinement as a form of reinforcement learning without weight updates.
After each trial, the agent reflects on what went wrong by generating a
natural-language critique (``The function failed because it did not handle
negative inputs''). This reflection is stored in an episodic memory buffer and
retrieved in subsequent attempts, conditioning the model to avoid past
mistakes. Reflexion accepts diverse feedback: scalar rewards (e.g., pass rate),
binary verdicts, or free-form error messages. The key innovation is converting
feedback into \emph{verbal} form that persists across trials, enabling the
model to learn from experience within a single episode. On HumanEval, Reflexion
achieves 91\% pass@1, surpassing single-shot GPT-4 (80\%) by explicitly
encoding trial-and-error history. The approach generalises beyond code: it
improves sequential decision-making in games and language reasoning tasks.

\paragraph{Self-debugging: rubber duck debugging.}
Chen~et~al.~\cite{chen2024selfdebug} propose \emph{self-debugging}, where the
model critiques and repairs its own code without external feedback. The agent
generates code, then \emph{explains} the code line-by-line in natural language.
This explanation serves as a self-generated specification, allowing the model
to spot discrepancies between intent and implementation. If the explanation
reveals a flaw (``This loop iterates over indices, but I wanted to iterate over
values''), the model revises the code. Self-debugging also incorporates
external execution feedback when available: the model receives error messages
or failing test outputs and reasons about root causes before proposing fixes.
On Spider (text-to-SQL), self-debugging improves baseline accuracy by 2--3\%
without unit tests and by up to 12\% on MBPP when tests are available. The
method demonstrates that internal self-critique complements external execution
signals.

\paragraph{Repair prompts and diff-based editing.}
Simpler refinement strategies use \emph{repair prompts}: the agent receives the
original task, the buggy code, and the error message, then generates a revised
version. To reduce token costs, some systems use \emph{diff-based editing},
where the model outputs only the changes (insertions, deletions) rather than
the full code. This is particularly effective for large files: instead of
regenerating hundreds of lines, the agent specifies ``replace line 42'' or
``insert after line 57''. SWE-agent's edit tool and Agentless's diff-based
repair both adopt this paradigm. The trade-off is that models must learn the
diff format, and parsing errors can break the repair loop.

\subsection{Does Refinement Work? The Limits of Self-Correction}
\label{sec:refinement-limits}

The results above are usually reported as evidence that agents improve their own
code. Two lines of work complicate that reading, and taking them seriously
sharpens rather than weakens the control-layer thesis.

\paragraph{Self-repair is partly inference-time scaling in disguise.}
Olausson et al.~\cite{olausson2024selfrepair} observe that a self-repair loop
does not spend a fixed budget: producing a critique and a revision consumes
tokens and samples that could instead have been spent drawing more independent
candidates from the generator. When they normalise for that budget, comparing
self-repair against simply sampling $k$ programs i.i.d.\ and taking the best,
much of self-repair's apparent gain disappears. The gain that survives is
attributable to the \emph{feedback}, not to the iteration: repair helps most
when the model receives a stronger critique than it can produce for itself, and
in their strongest condition the critique comes from a human. Reflexion's
HumanEval headline (\S\ref{sec:refinement-mechanisms}) should therefore be read
as a result about a \emph{budget}, not only about a mechanism, and compared
against a best-of-$N$ baseline at matched cost rather than against single-shot
GPT-4.

\paragraph{Intrinsic self-correction does not reliably help.}
Huang et al.~\cite{huang2024selfcorrect} isolate the case where no external
signal is available, which they term \emph{intrinsic} self-correction, and find
that models asked to review and revise their own reasoning often make it worse,
converting correct answers into incorrect ones roughly as often as the reverse.
The mechanism is unsurprising given \S\ref{sec:feedback-signals}: the critic and
the generator are the same distribution, so the critique cannot introduce
information the generator lacked. Reported gains from ``self-critique'' commonly
depend on an oracle that has quietly leaked into the loop, most often a stopping
rule that halts only once the answer is right.

\paragraph{What this implies for the pipeline.}
The two results converge on one conclusion, which we take as this report's
sharpest claim about refinement. \emph{Refinement is a control layer only when an
external oracle closes the loop; internal self-critique is an open loop wearing a
control loop's clothes.} The framing explains the pattern in
Table~\ref{tab:loop-closure}: every mechanism that survives contact with
repository-scale evaluation (Agentless's validation phase, SWE-agent's
edit--execute loop, D4C's test-driven repair) is externally closed, and the
purely internal mechanisms cluster in function-level benchmarks where a
comprehensive test suite is nonetheless available to the \emph{evaluator}, if
not to the agent.

It also unifies two techniques that the literature treats separately. Iterative
self-repair and best-of-$N$ sampling with a reranker are the same lever, a
budget of extra inference spent to reduce variance, pulled in two different
directions: sequentially, conditioning each sample on the last, or in parallel,
conditioning none of them. Once budget is held fixed the interesting question is
not ``does the agent refine?'' but ``which allocation of the same tokens buys
more correctness?'' OpenHands's Verified result answers it in one direction:
five parallel rollouts reranked by a dedicated critic reach 66.4\% where a single
rollout reaches 60.6\%~\cite{openhands2025verified}. Notably, the critic is a
\emph{separate, trained} model rather than the generator judging itself, which is
precisely the condition Olausson et al.\ identify as the one under which
feedback helps. The cost of this lever, and its consequences, return in
\S\ref{sec:open}.

\subsection{Fault Localisation and Self-Repair}
\label{sec:fault-localization}

When refinement targets bugs in existing code rather than improving draft
solutions, it becomes \emph{automated program repair} (APR). The classic APR
pipeline has three stages: fault localisation (identify buggy lines), patch
generation (propose fixes), and patch validation (test the fix). LLM-based
agents challenge this structure by generating end-to-end repairs.

\paragraph{Traditional fault localisation.}
Pre-LLM APR systems rely on \emph{spectrum-based fault localisation} (SBFL),
which ranks code lines by suspiciousness scores derived from test coverage:
lines executed by failing tests and not by passing tests receive high scores.
Ochiai, Tarantula, and DStar are standard SBFL metrics. LLM-based systems
initially adopted this pattern, using SBFL to extract a small code snippet and
prompting the model to repair it. However, fault localisation is imperfect:
even oracle-provided bug locations (``perfect FL'') do not guarantee correct
repairs, and SBFL scores are often noisy.

\paragraph{FL-free repair with objective alignment.}
Recent work questions whether explicit fault localisation is necessary.
D4C~\cite{xu2024d4c} argues that LLMs should
repair entire functions rather than isolated hunks, aligning the task with
their pre-training objective (next-token or infilling completion). D4C prompts
the model with the full function, failing test cases, and error messages, and
asks it to output a corrected function. On Defects4J, D4C repairs 180
single-function bugs with only 10 samples per bug, outperforming systems with
perfect fault localisation by 10\% and reducing sample count by 90\%. The
insight is that LLMs, trained on vast corpora of correct code, implicitly learn
to locate faults when given sufficient context; forcing them to target specific
lines may constrain their reasoning.

\paragraph{Reconciling FL-free repair with localisation-first agents.}
Read beside \S\ref{sec:planning}, D4C appears to contradict the systems surveyed
there. Agentless and AutoCodeRover treat localisation as the core of the problem
and spend most of their pipeline on it; D4C reports that localisation is
unnecessary and even harmful. Both are right, because they operate in different
regimes, and naming the regime is more useful than picking a winner.

D4C's Defects4J tasks are \emph{single-function} bugs: the buggy function is
supplied. What D4C eliminates is therefore not the search for the faulty
function but the finer-grained ranking of \emph{lines within} a function already
known to be faulty, and its argument, that line-level targeting fights the
model's infilling prior, is convincing at that granularity. A SWE-bench agent
faces a prior question, which function among tens of thousands, and no
pre-training prior answers it: the relevant symbol may not appear in the issue
text at all. The reconciliation is a claim about scale. \emph{Below} the
function boundary, localisation is a constraint the model does not need and
should not be given; \emph{above} it, localisation is the task. This is why
AutoCodeRover's SBFL variant helps (it ranks functions) while D4C's ablation of
perfect FL helps (it un-ranks lines), and the two results, superficially
opposed, are measuring opposite sides of the same boundary.

\paragraph{Stack-trace analysis and root-cause reasoning.}
When execution fails, agents receive stack traces showing the call chain and
exception type. Stack traces are rich signals: they identify the failing
function, reveal the exception (NullPointerException, IndexError), and often
include variable values. Agents can parse stack traces to infer root causes:
``IndexError at line~15 suggests the list is shorter than expected; check loop
bounds.'' SelfEvolve~\cite{jiang2023selfevolve} and SWE-agent~\cite{yang2024sweagent}
iteratively refine code by feeding error messages back to the LLM, which
reasons about the fault and proposes targeted edits. The challenge is that stack traces can be
noisy (multiple exceptions, deep call stacks) and may not reveal semantic bugs
(e.g., off-by-one errors that do not crash).

\paragraph{Test-driven repair loops.}
A refinement loop driven by test feedback proceeds as follows: (1)~generate
initial code, (2)~run tests, (3)~if any fail, extract failure messages,
(4)~prompt the model to fix the code, (5)~repeat until all tests pass or a
budget is exhausted. The stopping condition is critical: systems may iterate 1,
3, 5, or 10 times. Empirically, most gains occur in the first 2--3 iterations;
further iterations yield diminishing returns and risk introducing new bugs.
SWE-agent's edit--execute loop and Agentless's repair--validate pipeline both
iterate until tests pass or a budget is exhausted. The iteration budget
balances quality and cost.

\paragraph{Candidate ranking and ensemble repair.}
When multiple repair attempts produce different patches, agents must select the
best one. Ranking strategies include: (i)~test coverage (choose the patch
that passes the most tests), (ii)~LLM-as-judge (ask the model to explain and
rank patches by confidence), (iii)~ensemble voting (generate $N$ patches and
pick the most common or most syntactically similar to the original), and
(iv)~semantic similarity (prefer patches that minimally alter the original
code). CodeT~\cite{chen2022codet} and Agentless~\cite{xia2024agentless} both
rank candidates by test execution, filtering implausible patches before final
selection.

\subsection{Convergence and Stopping Criteria}
\label{sec:convergence}

Iterative refinement must terminate. Agents need stopping criteria that balance
solution quality, cost, and the risk of overfitting to weak test suites.

\paragraph{Pass-rate thresholds.}
The simplest criterion is \emph{full test pass}: stop when all provided tests
pass. This works well for competitive programming (HumanEval, CodeContests)
where test suites are comprehensive. However, in software engineering tasks
(SWE-bench), tests may be incomplete or incorrect. A patch that passes all
tests may still be wrong (a ``plausible but incorrect'' patch), and a correct
patch may fail tests due to environment issues. Note that the criterion is
also, quietly, an oracle: as \S\ref{sec:refinement-limits} observed, a loop
that halts exactly when the answer becomes right has smuggled ground truth into
its stopping rule, and any self-critique operating inside such a loop will
appear more effective than it is.

\paragraph{Diminishing improvement detection.}
If successive iterations yield no improvement (measured by test pass rate,
execution success, or the failing test remaining unchanged), the agent should
stop. If iterations 3, 4, and 5 all produce code that fails the same test with
the same error, further iteration is unlikely to help, and the heuristic
prevents endless loops in which the model oscillates between two incorrect
states. The empirical basis for stopping early is the shape of the repair curve:
Olausson et al.\ find that the marginal return of an additional repair round
falls off quickly and, once the token budget is held fixed, is frequently
negative relative to drawing a fresh independent
sample~\cite{olausson2024selfrepair}. Where the loop is closed only internally,
Huang et al.\ find the marginal return can be negative in absolute
terms~\cite{huang2024selfcorrect}: iterating past the first round makes the
answer worse. The practical implication is that a stopping rule and a budget
allocation are the same decision, and that ``iterate until it passes'' is
defensible only when an external oracle defines \emph{passes}.

\paragraph{Budget limits: iterations and tokens.}
Practical systems impose hard limits. Iteration budgets (3, 5, 10 rounds) cap
the number of refinement cycles. Token budgets limit total inference cost:
after consuming a threshold, the agent returns its best
attempt. Reflexion and Agentless both use iteration caps, while D4C's low
sample count (10 tries per bug) implicitly limits token use. The optimal budget
depends on task difficulty: simple bugs may resolve in one iteration, while
complex logical errors may require many. The budget is not merely a cost
control, however. Because sequential refinement and parallel best-of-$N$ draw on
the same pool of tokens (\S\ref{sec:refinement-limits}), fixing the budget is
what makes the two strategies comparable at all, and an iteration cap is
implicitly a bet about which of them converts tokens into correctness more
efficiently for the task at hand.

\paragraph{Risk of overfitting to weak test suites.}
A critical failure mode is overfitting. If the test suite is weak (few tests,
shallow coverage), the agent may generate code that narrowly passes those tests
but fails on unseen inputs. This is the ``plausible patch'' problem in APR:
patches that satisfy the validation oracle but do not generalise. Mitigations
include: (i)~\emph{test augmentation}, where the agent or a separate test
generator produces additional test cases to catch edge cases; (ii)~\emph{static
checks}, requiring generated code to satisfy linting, type checking, or
invariant assertions beyond test pass; (iii)~\emph{semantic similarity
constraints}, penalising patches that radically alter program structure; and
(iv)~\emph{LLM-based verification}, asking the model to critique the patch or
predict whether it will generalise. However, no technique fully eliminates
overfitting, and the problem remains open.

\paragraph{Human-in-the-loop approval.}
In high-stakes domains (medical software, financial systems), automated
iteration may be unacceptable without human oversight. Systems like OpenHands
and Devin allow developers to inspect intermediate outputs, approve or reject
patches, and provide guidance before the next iteration. This trades speed for
safety and introduces latency, but ensures that convergence criteria align with
human judgment rather than proxy metrics.

\section{Discussion}
\label{sec:discussion}

Having surveyed end-to-end systems (\S\ref{sec:e2e}), planning and search
(\S\ref{sec:planning}), and execution-feedback refinement
(\S\ref{sec:refinement}), we now consolidate what is common across these
fronts and which cross-cutting constraints bound all of them. Progress has come
less from any single architectural idea than from \emph{coupling} generation to
external evidence, and that same coupling now exposes the field's hardest
problems: evaluation integrity, finite context, cost, and trust.

\subsection{Consolidated Patterns and Trends}
Three design patterns recur across nearly every system we surveyed, and one
consequence follows from all three.

The \emph{observe--think--act loop}~\cite{yao2023react} is the near-universal
control structure (\S\ref{sec:e2e-arch}); Agentless's fixed
pipeline~\cite{xia2024agentless} is the exception that stays competitive by
hard-coding the sequence the loop would otherwise discover. \emph{Execution-grounded
selection}, which ranks sampled candidates by tests, generated tests, or
regression preservation rather than by model confidence
(AlphaCode~\cite{li2022alphacode}, CodeT~\cite{chen2022codet}, Agentless), is
the most reliable quality lever. And \emph{planning functions as a control layer
over generation} rather than a standalone up-front step (\S\ref{sec:planning}).

The unifying observation is that the second pattern explains the other two.
Everything that works, works by putting something outside the model in a
position to contradict it: a test runner, a compiler, an AST index, a trained
critic. Planning earns its place by grounding hypotheses in repository structure
that can be checked (\S\ref{sec:planning}); refinement earns its place only when
an external oracle closes the loop, and degrades to noise when it does not
(\S\ref{sec:refinement-limits}). Read this way the pipeline is not five stages of
reasoning but a scaffold for arranging confrontations between a generative model
and an environment that can say no. Its two weakest points are exactly the two
places where nothing can: intent capture, where no oracle exists at all
(\S\ref{sec:intent}), and test-based validation, where the oracle exists but is
weak enough to be gamed (\S\ref{sec:eval-landscape}).

This suggests a sharper way to state the control-layer thesis. A control loop
regulates a plant against a \emph{setpoint}; for these agents the setpoint is
``the tests pass.'' But \S\ref{sec:convergence} and \S\ref{sec:open} together
establish that this setpoint is a proxy, and a leaky one: patches that pass can
be overfitted, insecure, or the product of reward hacking. Current agents are
therefore control loops regulating tightly on a signal that only approximates
what we want. Their impressive convergence and their characteristic failures are
the same phenomenon viewed from either end.

Three macro trends follow. The field has moved from single-shot prompting to
agentic scaffolds; from flat chain-of-thought to structured search and
localisation; and, most recently, from a one-shot rollout to
\emph{inference-time scaling}, where many trajectories are sampled and reranked
by a critic model, as OpenHands's 60.6\%~$\to$~66.4\% Verified result
illustrates~\cite{openhands2025verified}. A consistent
finding across \S\ref{sec:e2e} is that \emph{scaffold and interface matter as
much as the backbone model}. That finding, however, is time-stamped rather than
timeless. A newer line of work moves the scaffold's competence into the weights,
training agents on execution feedback directly, either by reinforcement learning
over software-evolution data~\cite{wei2025swerl} or by training agent and
verifier together in an executable environment~\cite{pan2024swegym}. If this
succeeds, the scaffolding advantage that \S\ref{sec:e2e} documents is a
transitional artefact of using models that were never trained to be agents.
A parallel shift has occurred in deployment: the systems surveyed here are
research artefacts, while the agents developers actually use (Claude Code,
Cursor, OpenAI's Codex) are products whose scaffolds are undocumented and whose
evaluations are self-reported, which is one reason the benchmark integrity
questions of \S\ref{sec:eval-landscape} have become urgent.

\subsection{The Evaluation Landscape}
\label{sec:eval-landscape}
The benchmarks used to measure agentic code generation form a layered stack,
and each layer is showing strain. \emph{Function-level} suites (HumanEval, MBPP)
are effectively saturated (Reflexion already reported 91\% pass@1 on HumanEval
in 2023~\cite{shinn2023reflexion}) and no longer discriminate among frontier
systems. \emph{Repository-level} evaluation shifted the centre of gravity to
SWE-bench~\cite{jimenez2024swebench}, whose Verified split~\cite{openai2024verified}
has become the de-facto leaderboard standard. Yet \S\ref{sec:e2e-eval} already
noted its difficulty skew, which in numbers means 39\% of tasks are trivial
$<$15-min fixes and 86\% touch a single file~\cite{ganhotra2025difficulty}, its
repository monoculture (Django dominates), and its contamination risk, since many
issues predate model training cutoffs and the repositories are widely
scraped~\cite{epoch2025swebench}.

Three responses have emerged.
The first is \emph{harder, contamination-resistant successors}. SWE-bench Pro~\cite{deng2025swebenchpro}
contributes 1{,}865 long-horizon problems from 41 repositories and resists
memorisation using copyleft-licensed and held-out proprietary code; at release,
top systems scored only $\sim$23\% versus 70\%+ on Verified, and even by
mid-2026 no public system clears the high~50s~\cite{agentmarketcap2026}.
SWE-bench Multilingual~\cite{swebench2025multilingual} extends the format to 300
tasks across 9 languages, addressing SWE-bench's Python monoculture.
Figure~\ref{fig:verified-vs-pro} makes the divergence concrete: frontier models
clustered near or above 80\% on Verified fall to the mid-40s--high-50s on Pro, a
20--35 point gap for the \emph{same} model weights. Claude Opus~4.5, for
instance, scores 80.9\% on Verified but 45.9\% on Pro under Scale's standardised
harness~\cite{deng2025swebenchpro,agentmarketcap2026gap}. Two confounds prevent
reading the \emph{size} of each model's gap as a memorisation measurement, and
we flag them because they are the same confounds this report holds others to.
Pro is harder as well as fresher, so difficulty and contamination are entangled
in every drop; and the Pro column mixes harnesses, only Opus~4.5 being scored
under Scale's standardised evaluation while the rest self-report with custom
scaffolding. The systematic gap across all four models is robust to both. The
per-model ranking within it is not, and we draw no conclusion from it. The signal was strong
enough that OpenAI stopped reporting Verified scores in early 2026, citing
near-universal contamination (models reproduce gold patches from the task ID
alone) and flawed residual tests~\cite{openai2026swebench}. The second
response targets \emph{benchmark integrity directly}: a stricter evaluation
harness that closes test-leakage channels can cut reported scores substantially
(e.g., double-digit drops on SWE-bench Pro for some models), prompting the
observation that ``reward hacking is swamping model intelligence gains'': agents
partly learn to exploit weaknesses in the grading oracle rather than to solve
the task~\cite{cursor2025rewardhacking}. The third response is
\emph{efficiency-aware evaluation}: SWE-Effi~\cite{sweeffi2025} re-scores agents
under resource constraints and finds that resolve rate decouples sharply from
cost. Agentless reaches the highest resolve rate in its study (48\% with
Qwen3-32B) but consumes 83.1 API calls and 209.4 seconds per issue on average,
so a single pass@1 number conceals an order-of-magnitude spread in cost.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      width=\columnwidth,
      height=5.2cm,
      ybar,
      bar width=5pt,
      enlarge x limits=0.18,
      ymin=0, ymax=100,
      ylabel={\% Resolved},
      symbolic x coords={Opus 4.5, GPT-5.2, GPT-5.3, GLM-5.1},
      xtick=data,
      x tick label style={font=\scriptsize, rotate=18, anchor=east},
      yticklabel style={font=\scriptsize},
      ylabel style={font=\small},
      legend style={font=\tiny, at={(0.5,1.02)}, anchor=south,
                    legend columns=2, draw=none, fill=none, /tikz/every even column/.append style={column sep=6pt}},
      grid=major, grid style={dashed, gray!30},
      nodes near coords, point meta=explicit symbolic,
      every node near coord/.append style={font=\tiny, rotate=90, anchor=west},
    ]
    \addplot[fill=blue!55!black, draw=blue!55!black] coordinates {
      (Opus 4.5,80.9)  [80.9]
      (GPT-5.2,82)     [82]
      (GPT-5.3,85)     [85]
      (GLM-5.1,78)     [78]
    };
    \addplot[fill=orange!80!black, draw=orange!80!black] coordinates {
      (Opus 4.5,45.9)  [45.9]
      (GPT-5.2,56.4)   [56.4]
      (GPT-5.3,56.8)   [56.8]
      (GLM-5.1,58.4)   [58.4]
    };
    \legend{SWE-bench Verified, SWE-bench Pro}
    \end{axis}
  \end{tikzpicture}
  \caption{The Verified--Pro gap. The \emph{same} frontier model
    weights score 20--35 points lower on the harder, contamination-resistant
    SWE-bench Pro (1{,}865 tasks, 41 repos, GPL+held-out code) than on the
    saturated, widely-scraped SWE-bench Verified (500 tasks, 12 repos). Only the
    \emph{systematic} gap should be read from this figure. The per-model
    differences must not be: Verified scores are vendor-reported and
    approximate, and on Pro only Opus~4.5 is scored under Scale's standardised
    SEAL harness, the rest being self-reported with custom scaffolding. The
    model measured under the strictest harness will show the largest drop for
    reasons that have nothing to do with memorisation, so the spread across
    models here is confounded with the spread across harnesses. Data compiled
    April~2026 from leaderboard
    aggregators~\cite{agentmarketcap2026gap,deng2025swebenchpro}.}
  \label{fig:verified-vs-pro}
\end{figure}

A single resolve-rate number is therefore no longer a
sufficient summary. Credible comparison now requires difficulty stratification,
contamination controls, and cost (or token/latency) normalisation reported
alongside accuracy. The pipeline framing makes the coverage gaps concrete: most
benchmarks still score only the \emph{end} of the pipeline (did the final patch
pass?), leaving intermediate stages such as plan quality, localisation
precision, and refinement efficiency largely unmeasured.

\subsection{The Context Window Bottleneck}
Every pipeline stage is constrained by the same resource: the model's finite
working context. Real repositories far exceed even million-token windows, so an
agent never sees the whole program and must decide \emph{what} to read. This is
not solved merely by longer context windows. Repository-level studies find that
retaining the full \emph{dependency} context yields the best results, while
simply enlarging the context with loosely related code can be
misleading and degrade correctness~\cite{repoexec2025}: more tokens are not
uniformly better. A second failure mode is dynamic: agents that append every
observation to an ever-growing trajectory suffer \emph{context explosion} and
\emph{semantic drift}, where critical early information is buried or lost as the
conversation lengthens~\cite{contextastool2025}. Because input tokens are
re-sent each turn, trajectory length also drives cost roughly quadratically over
a run~\cite{turncontrol2025}, tightly coupling the context bottleneck to the
cost challenge below.

The systems we surveyed therefore treat context as a managed resource rather
than a passive log. Mitigations span the whole pipeline: hierarchical
repository skeletons with embedding retrieval (Agentless~\cite{xia2024agentless}),
AST-based structure-aware search that returns signatures instead of full bodies
(AutoCodeRover~\cite{zhang2024autocoderover}), and observation compression of
older turns (SWE-agent~\cite{yang2024sweagent}). The current research direction
is to make context management \emph{active}, explicitly summarising, pruning,
and re-retrieving context as a deliberate agent action rather than a passive
heuristic~\cite{contextastool2025}. Context quality, not model capability
alone, frequently determines whether the correct patch is even reachable: an
agent cannot fix what it never places in context.

\subsection{Open Challenges}
\label{sec:open}
Four challenges cut across all three fronts and remain open.

\paragraph{Long-horizon, multi-file reasoning.}
Current agents excel at small, local, single-file fixes but degrade steeply on
tasks that require coordinated changes across many files and long action
horizons. The easy/hard gap on SWE-bench Verified ($\sim$80\% of easy vs.\
$\sim$20\% of hard tasks solved by the best single
system~\cite{ganhotra2025difficulty}) and the steep drop on the explicitly
long-horizon SWE-bench Pro~\cite{deng2025swebenchpro} both point to the same
ceiling. Sustained planning, memory, and error recovery over long trajectories
remain unsolved.

Is this the same constraint as the context bottleneck above, or a distinct one?
The evidence in \S\ref{sec:e2e-eval} suggests they are distinct, and we commit
to that reading. If the hard tier were context-bound, it would be an
\emph{ensembling} problem, since different scaffolds retrieve different context
and would therefore fail on different instances. That is exactly what the easy
and medium tiers show: pooled across nine systems, some agent solves 95.4\% of
easy and 84.3\% of medium instances, far above any single system. The hard tier
does not behave this way. More than half of it defeats all nine systems at once,
and only 9 of 25 hard multi-file issues are solved by
anyone~\cite{ganhotra2025difficulty}. A limitation that every scaffold hits at
the same instances is not a retrieval limitation. Context management is
necessary and, on the evidence, insufficient: it determines whether the right
patch is \emph{reachable}, while the hard tier appears to test whether the model
can construct a multi-step, cross-file change that no amount of retrieval will
hand it. We would expect training-time
approaches~\cite{wei2025swerl,pan2024swegym} rather than better retrieval to
move this tier, and offer that as a falsifiable prediction.

\paragraph{Specification ambiguity.}
This is the open-loop intent stage of \S\ref{sec:intent}, seen from the outside.
Vibe coding starts from natural-language intent that is often
underspecified. When the issue text omits edge cases or implicit invariants, the
agent must guess, and execution feedback only constrains behaviour that the
available tests exercise, so a wrong guess about an untested invariant is never
contradicted and is scored as success. Closing the gap between an ambiguous
intent and a verified specification, whether through clarification, test
synthesis, or interactive intent capture, is largely unaddressed in current
end-to-end systems, and current benchmarks cannot reward addressing it: a
dataset built from issues that a real pull request resolved is a dataset from
which ambiguity has been filtered out in advance.

\paragraph{Cost and efficiency.}
The dominant quality lever, inference-time scaling with many sampled
trajectories and critic reranking, directly multiplies cost, while trajectory
length inflates token usage per turn~\cite{turncontrol2025}. Because resolve
rate decouples from compute~\cite{sweeffi2025}, the practical frontier is not
peak accuracy but accuracy per dollar and per minute. Turn-control and
trajectory-reduction strategies are early attempts to make agents economically
deployable rather than merely capable.

\paragraph{Trustworthiness and security.}
Passing the available tests does not make generated code safe. A large-scale
evaluation across 100+ models found that 45\% of generation tasks introduced a
known security flaw, and that this rate has not fallen as models have grown
larger and newer~\cite{veracode2025genai}. Purpose-built benchmarks for
agentic settings reach the same conclusion: SusVibes shows that vibe-coding
agents produce vulnerable implementations on realistic feature
requests~\cite{susvibes2025}, and systematic analyses document insecure
behaviours that functional test suites never catch~\cite{securitydebt2025}.
Because refinement loops optimise for \emph{functional} pass rates
(\S\ref{sec:convergence}), they do not penalise insecure-but-passing patches, a
blind spot that compounds with the overfitting risk of weak test suites.

This is where the terminology of \S\ref{sec:vibe-definition} comes due. In
Karpathy's narrow sense, vibe coding is defined by the developer not reading the
diff. We argued that the pipeline must then re-supply every guarantee the reader
used to supply. It does re-supply some: localisation replaces the developer's
knowledge of where the bug lives, and the test runner replaces their judgement
of whether it is fixed. But no stage re-supplies the security review, because no
stage has an oracle for it, and the 45\% figure is what that missing oracle
costs. Human-in-the-loop review (OpenHands, Devin) remains the default
safeguard, but it reintroduces exactly the reading that vibe coding was defined
by abandoning. The honest statement of the situation is that vibe coding, in the
strict sense, is not yet a supported mode of operation; it is a mode of
operation in which one has decided not to look.

\subsection{Implications for Software Engineering Practice}
\label{sec:practice}
These trends shift the developer's role rather than remove it. As agents absorb
localisation, editing, and test-driven repair, the human contribution moves
upstream (writing precise intent and acceptance criteria) and downstream
(reviewing, verifying, and accepting patches). Two consequences follow from the
evidence above. First, \emph{verification becomes the bottleneck}: with a known
security flaw in 45\% of generation tasks~\cite{veracode2025genai} and benchmark
scores partly reflecting reward hacking~\cite{cursor2025rewardhacking},
organisations cannot treat a green test run as acceptance, and must invest in
stronger oracles (security scanning, regression suites, and human review) as
first-class pipeline stages. Second, \emph{execution infrastructure becomes
central}: sandboxes, reproducible environments, and CI integration are what make
execution-grounded selection and refinement possible in the first place. The
economic frame is shifting too, from ``can an agent resolve this issue?'' to
``at what cost, and with what assurance?''

\section{Conclusion}
This report surveyed autonomous AI agents for code generation through a single
lens, the five-stage vibe coding pipeline (intent, planning, generation,
execution, refinement), and synthesised three active research fronts that
instantiate its stages. From end-to-end systems (\S\ref{sec:e2e}) we found a
convergent architecture: an observe--think--act loop over a purpose-built
agent--computer interface, with repository-scale context managed by retrieval,
AST search, or compression. Agentless shows that a fixed localise--repair--validate
pipeline can rival interactive agents at lower cost, so autonomy is a design
choice, not a prerequisite. From planning and search (\S\ref{sec:planning}) we
found that what planning contributes is not deliberation but \emph{grounding}:
localisation hypotheses and candidate search anchored in program structure,
which is why the least deliberative system remains competitive. From execution
feedback (\S\ref{sec:refinement}) we found that running the code, not generating
it, supplies most of the reliability, and that knowing \emph{when to stop} is as
important as knowing how to fix.

The pipeline framing makes the field's central coupling visible: planning and
refinement, usually studied in isolation, together form a feedback control layer
that keeps generation tethered to external evidence. The framing also sharpens
the claim to something falsifiable. That control layer exists only where an
external oracle can contradict the model; internal self-critique, once
budget-normalised, is an open loop wearing a control loop's clothes
(\S\ref{sec:refinement-limits}). Read this way, the pipeline's two structural
weaknesses are the two stages with no oracle behind them: intent capture, which
has none at all (\S\ref{sec:intent}), and validation, whose oracle is weak enough
to be gamed (\S\ref{sec:eval-landscape}). The same coupling
exposes the limits surveyed in \S\ref{sec:eval-landscape}--\S\ref{sec:open}: a
saturating and contamination-prone evaluation landscape, a persistent context
bottleneck, escalating cost, and a trust gap in which 45\% of generation tasks
yield code with a known security flaw~\cite{veracode2025genai}. The question at
the frontier has changed accordingly. It is no longer ``can an agent resolve
real issues?'' but ``at what cost, and how much can the result be trusted?'',
and contamination-resistant benchmarks such as SWE-bench
Pro~\cite{deng2025swebenchpro}, efficiency-aware scoring~\cite{sweeffi2025}, and
reward-hacking audits~\cite{cursor2025rewardhacking} are the instruments the
field built to answer it.

The promising directions follow from that, and each amounts to supplying an
oracle where one is missing. On evaluation, the community needs harder,
multilingual, contamination-resistant benchmarks reported with cost and
difficulty stratification, and benchmarks that score the pipeline's intermediate
stages rather than only its final patch. On capability, long-horizon multi-file
reasoning appears to be a training problem rather than a retrieval one
(\S\ref{sec:open}), and active context management the binding constraint on
reachability. On specification, an agent that detects its own underspecified
intent and asks, rather than guessing, would close the one stage that no system
currently closes. On deployment, security-aware refinement objectives and
well-designed human-in-the-loop control are needed before autonomous code
generation can be trusted in production. Vibe coding has clearly crossed from
autocomplete to autonomous issue resolution. Whether it crosses from impressive
to dependable will be decided by whether the pipeline learns to say no to itself
as reliably as a reader once did.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
